---
title: "SuicideRate_Final"
author: "Aditi Dutt, Avijeet Bose, Sangeeta Singh, Christian Candler"
date: "11/21/2020"
output:
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
## Libraries
library(data.table)
library(forecast)
library(leaps)
library(MASS)
library(gplots)
library(tidyverse)
library(ggplot2)
library(GGally)
library(scales)
library(mosaic)
library(mapproj)
library(mlbench)
library(caret)
library(ggpubr)
library(gridExtra)
library(pivottabler)
library(rpivotTable)
library(dplyr)
library(reshape2)
library(tinytex)
#library(ggalt) # dumbbell plots
library(countrycode) # continent
library(rworldmap) # quick country-level heat maps
library(gridExtra) # plots
library(broom) # significant trends within countries

if(!require(tidyverse)) install.packages("tidyverse", repos = "http://cran.us.r-project.org")
if(!require(caret)) install.packages("caret", repos = "http://cran.us.r-project.org")
if(!require(data.table)) install.packages("data.table", repos = "http://cran.us.r-project.org")
if(!require(ggcorrplot)) install.packages("ggcorrplot", repos = "http://cran.us.r-project.org")
if(!require(scales)) install.packages("scales", repos = "http://cran.us.r-project.org")
options(scipen=10000)
if(!require(ranger)) install.packages("ranger", repos = "http://cran.us.r-project.org")
if(!require(countrycode)) install.packages("countrycode", repos = "http://cran.us.r-project.org")
options(scipen=10000)
```

```{r}
suiciderate_data.df <- read.csv("suiciderate.csv")

# Selecting variables
suiciderate_data_select.df <- suiciderate_data.df
suiciderate_data_select.dt <-data.table(suiciderate_data_select.df)
suiciderate_data_select.dt
```

```{r}
# Renaming
suiciderate_newdata.df <- rename(suiciderate_data.df, c(Country  = "ï..country", Year = "year", Sex = "sex", Age = "age", No_of_Suicides = "suicides_no", Population = "population", Suicide_Per_Population = "suicides.100k.pop", GDP = "gdp_for_year....", GDP.per.capita = "gdp_per_capita....", Generation = "generation"))
```

```{r}
minimum_years <- suiciderate_newdata.df %>% # using suicide_data
  group_by(Country) %>% # group all the data from the same country
  summarise(rows = n(), years = rows / 12, .groups = 'drop') %>% # collate number of rows 
  # for each country, with second column denoting the number of years
  # worth of data for each country
  arrange(years) # arrange the table by years (ascending)

suiciderate_newdata.df <- suiciderate_newdata.df %>% # using suicide_data
  filter(!(Country %in% head(minimum_years$Country, 10))) # remove the 
  # countries that are in the first 10 rows of the minimum_years table
```

```{r}
# Making age ordinal
suiciderate_newdata.df$Age <- factor(suiciderate_newdata.df$Age, ordered =T,
                                     levels = c('5-14 years',
                                                '15-24 years',
                                                '25-34 years',
                                                '35-54 years',
                                                '55-74 years',
                                                '75+ years'))
```

```{r}
# Making generation ordinal
suiciderate_newdata.df$Generation <- factor(suiciderate_newdata.df$Generation, 
                   ordered = T, 
                   levels = c("G.I. Generation", 
                              "Silent",
                              "Boomers", 
                              "Generation X", 
                              "Millenials", 
                              "Generation Z"))
```

```{r}
# checking enough data available for each year
years <- suiciderate_newdata.df %>%
  group_by(Year) %>%
  summarise(rows = n(), .groups = 'drop')
```

```{r}
plot_years_theme <- theme(axis.title = element_text(size = 15), 
                          axis.text = element_text(size = 10))
plot_years <- ggplot(data = years, 
                     aes(x = Year, y = rows, 
                         fill = factor(ifelse(Year == '2016',
                                              'Highlighted',
                                              'Normal')))) +
  # fill ifelse separates 2016 and allows it to be coloured red
  geom_bar(stat = 'identity', show.legend = F) + # create bar chart,
  # remove legend
  theme_classic() + # white background, no gridlines
  xlab('Year') + # change x axis label
  ylab('Rows of Data') + # change y axis label
  plot_years_theme + # change the size of axis titles and axis text
  scale_x_continuous(breaks = c(seq(1985, 2016, 2)),
                     labels = c(seq(1985, 2016, 2)),
                     limits = c(1984, 2017)) +
  scale_y_continuous(breaks = c(seq(0, 1250, 250)),
                     labels = c(seq(0, 1250, 250)),
                     limits = c(0, 1250)) +
  # change x and y axis values
  scale_fill_manual(name = 'Year', values = c('red', 'black'))
  # fill every bar black, except in 2016 which is red
plot_years
```
It seems that not all years included in the dataset have the same number of observations. The following plot shows that the year 2016 has the least number of observations. For this reason, year 2016 will be excluded later in the analysis

```{r}
# thus we remove year 2016
suiciderate_newdata.df <- subset(suiciderate_newdata.df, subset = Year < 2016)
```

## Exploratory Data Analysis

###Defining the Target Variable:`Suicide_Per_Population`

The distribution of the `Suicide_Per_Population` variable seems to be extremely positively skewed to the right. This shape suggests that the majority of the observations have a very small value of suicide rate with a small number of observations with very high values, causing a very long positive tail to the shape of the distribution. This severe skewness might suggest the use of a transformation. To handle this we are going to do the log transformation of the target variable before doing the regression test

```{r suicide_rate_histogram, echo=FALSE, fig.align="center", fig.height=3, fig.width=5}
# Distribution of Suicide_Per_Population
suiciderate_newdata.df %>% 
  ggplot(aes(Suicide_Per_Population)) +
  geom_histogram(fill="deepskyblue2", color="navy") +
  labs(y="Count", x="Suicide rate per 100k") +
  ggtitle("Distribution of Suicide Rates per 100K")
```
```{r}
# Variable transformation
suiciderate_newdata.df <- suiciderate_newdata.df %>%
  mutate(suicide_rate_log=log(1+Suicide_Per_Population))

```

```{r rates_by_country, echo=FALSE, fig.align="center", fig.width=5}
suiciderate_newdata.df %>% 
  group_by(Country) %>%
  summarize(country_suicide_rate=sum(No_of_Suicides)*100000/sum(Population), .groups='drop') %>%
  top_n(25) %>%
  ggplot(aes(reorder(Country, country_suicide_rate), country_suicide_rate)) +
  geom_bar(stat="identity", fill="deepskyblue2", color="navy") +
  coord_flip() +
  labs(x="Country", y="Suicide rate per 100k population") +
  ggtitle("Suicide Rates by Country")
```

### Exploring Suicide Rate Variability by Country and Continent
Suicide rate varies from one country to another. This is confirmed by the following bar chart. For easier interpretation, the figure shows the top 25 countries in terms of suicide rates. Lithuania is universally the top country in terms of suicide at a little over 40 suicides per 100k population. This rate is extremely high especially that Lithuania is not a big country like Russia, for example, which comes second after Lituania in the ranking. In fact, the average population size in Lithuania between 1985 and 2016 is `r suicide %>% filter(country=="Lithuania") %>% summarize(rate=mean(suicide_rate)) %>% round(.$rate)` compared to `r suicide %>% filter(country=="Russian Federation") %>% summarize(rate=mean(suicide_rate)) %>% round(.$rate)`. This disproportion in the case of Lithuania induces further exploration of the country populatio size.

```{r add_continent}
# Add continent variable
suiciderate_newdata.df$Continent <- countrycode(sourcevar = suiciderate_newdata.df[, "Country"],
                                 origin = "country.name",
                                 destination = "continent")
```
Continent variable is not present in the dataset but it can easily be added using the countrycode package.



```{r rates_by_continent, echo=FALSE, fig.align="center", fig.width=5}
suiciderate_newdata.df %>% 
  group_by(Country, Continent) %>%
  summarize(avg_suicide_rate=mean(Suicide_Per_Population), .groups='drop') %>%
  ggplot(aes(Continent, avg_suicide_rate)) +
  geom_boxplot( fill="deepskyblue2", color="navy") +
  # coord_flip() +
  labs(x="Continent", y="Suicide reate per 100k population") +
  ggtitle("Suicide rate by Continent")
```
The distribution of suicide rate varies from one continent to another. The highest suicide rates are in Europe and the distribution is severly skewed to the right. We see similar skeweness in Asia and the Americas as well with the presense of a few extreme values in each continent. 



```{r}
# look at how global suicides per 100k population have changed
# calculate the global mean
global_mean = (sum(as.numeric(suiciderate_newdata.df$No_of_Suicides)) / 
                 sum(as.numeric(suiciderate_newdata.df$Population))) * 100000
suicides_global = suiciderate_newdata.df %>% # using suicide_data
  group_by(Year) %>% # group all the data from the same year
  summarise(population = sum(Population), 
            suicides = sum(No_of_Suicides), 
            suicides_per_100k = (suicides / population) * 100000, .groups = 'drop')
  # create rows with the sum of the global population, sum of suicides,
  # and the suicides per 100k population, for each year

plot_global_theme <- theme(axis.title = element_text(size = 15), 
                           axis.text = element_text(size = 10))
plot_global <- ggplot(data = suicides_global, 
                      aes(x = Year, y = suicides_per_100k)) +
  geom_point(lwd = 2) +
  # add data as points
  geom_line(lwd = 1) +
  # join points together
  geom_hline(yintercept = global_mean, linetype = 2) +
  # add horizontal line denoting the global mean for the period
  theme_classic() + # white background, no gridlines
  xlab('Year') + # change x axis label
  ylab('Mean Global Suicides per 100k Population') + 
  # change y axis label
  plot_global_theme + # change the size of axis titles and axis text
  scale_x_continuous(breaks = c(seq(1985, 2015, 2)),
                     labels = c(seq(1985, 2015, 2)),
                     limits = c(1984, 2016)) +
  scale_y_continuous(breaks = c(seq(11, 16, 1)),
                     labels = c(seq(11, 16, 1)),
                     limits = c(11, 16))
  # change x and y axis values
plot_global
```
## The dashed line indicates the global mean for the time period: 13.27 deaths (per 100k, per year). Interpretation:
#Suicide rate was relatively low for the first four years, then increases substantially,
#Suicide rate increases steadily over from 1989-1992, then increases substantially again over the next three years to a peak of 15.3 in 1995,
#Since 1995, the suicide rate has generally fallen and may reach the early low level in the next decade.
#Decreased suicide rate per 100k steadily to around 11.7 in 2015 (around 25% decrease)

Suicide rates vary from one year to another. The following timeplot shows that before 1995, there was a global ascending trend of suicide. The opposite is true after 1995, as we see suicide rates decrease in a downward trend.

```{r}
sex_plot <- suiciderate_newdata.df %>%
  group_by(Sex) %>%
  summarize(suicide_per_100k = (sum(as.numeric(No_of_Suicides)) / sum(as.numeric(Population))) * 100000, .groups = 'drop') %>%
ggplot(aes(x = Sex, y = suicide_per_100k, fill = Sex)) + 
  geom_bar(stat = "identity") + 
  labs(title = "Global suicides (per 100k), by Sex",
       x = "Sex", 
       y = "Suicides per 100k") +
  theme(legend.position = "none") + 
  geom_text(aes(label = round(suicide_per_100k,3)), size = 2.5,vjust = -0.25) +
  scale_y_continuous(breaks = seq(0, 25), minor_breaks = F)

### with time
sex_time_plot <- suiciderate_newdata.df %>%
  group_by(Year, Sex) %>%
  summarize(suicide_per_100k = (sum(as.numeric(No_of_Suicides)) / sum(as.numeric(Population))) * 100000, .groups = 'drop') %>%
  ggplot(aes(x = Year, y = suicide_per_100k, col = factor(Sex))) + 
  facet_grid(Sex ~ ., scales = "free_y") + 
  geom_line() + 
  geom_point() + 
  labs(title = "Trends Over Time, by Sex", 
       x = "Year", 
       y = "Suicides per 100k", 
       color = "Sex") + 
  theme(legend.position = "none") + 
  scale_x_continuous(breaks = seq(1985, 2015, 5), minor_breaks = F)

grid.arrange(sex_plot, sex_time_plot, ncol = 2)
```
## ***Insights***

#Globally, the rate of suicide for men has been **~3.5x higher for men**
#Both **male & female suicide rates peaked in 1995**, declining since
#This ratio of 3.5 : 1 (male : female) has remained relatively constant since the mid 90's
#However, during the 80's this ratio was as low as 2.7 : 1 (male : female)

When looking at suicide rates by sex, the data shows that suicide is more prevalent among males than females, with universal suicide rate of almost 21 versus about 6 suicides per 100k male population.
Suicide rate trends across time varies by sex as well. The following plot shows that the suicide rates for females exhibit a universal descending trend across time while the trend for males fluctuates around 1995 which, resembles the  trend we saw before.

```{r}
age_plot <- suiciderate_newdata.df %>%
  group_by(Age) %>%
  summarize(suicide_per_100k = (sum(as.numeric(No_of_Suicides)) / sum(as.numeric(Population))) * 100000, .groups = 'drop') %>%
  ggplot(aes(x = Age, y = suicide_per_100k, fill = Age)) + 
  geom_bar(stat = "identity") + 
  labs(title = "Global suicides per 100k, by Age",
       x = "Age", 
       y = "Suicides per 100k") +
  plot_age_theme +
  theme(legend.position = "none", axis.text.x = element_text(size = 6)) + 
  scale_y_continuous(breaks = seq(0, 30, 1), minor_breaks = F)

### with time
age_time_plot <- suiciderate_newdata.df %>%
  group_by(Year, Age) %>%
  summarize(suicide_per_100k = (sum(as.numeric(No_of_Suicides)) / sum(as.numeric(Population))) * 100000, .groups = 'drop') %>%
  ggplot(aes(x = Year, y = suicide_per_100k, col = Age)) + 
  facet_grid(Age ~ ., scales = "free_y") + 
  geom_line() + 
  geom_point() + 
  labs(title = "Trends Over Time, by Age", 
       x = "Year", 
       y = "Suicides per 100k", 
       color = "Age") + 
  theme(legend.position = "none") + 
  scale_x_continuous(breaks = seq(1985, 2015, 5), minor_breaks = F)


grid.arrange(age_plot, age_time_plot, ncol = 2)
```
#***Insights***

#Globally, the **likelihood of suicide increases with age**
#Since 1995, suicide rate for everyone aged **>= 15** has been **linearly decreasing**
#The suicide rate of those aged **75+** has dropped by more than **50%** since 1990
#Suicide rate in the **'5-14'** category remains **roughly static** and small (< 1 per 100k per year)

## Population Characteristics Variability
Now after taking an overview on suicide variability by several country and nation level variables, we move to demographic variables that characterise the populations of these countries. First, we look at suicide variability by age goup. The following plot shows that suicide rates distribution varies from one age group to another. The highest suicide rates are found in individuals aged 75+ and the lowest suicide rates are found in individuals aged between 5-14.

```{r rates_gdp, echo=FALSE, fig.align="center", fig.height=3, fig.width=5}
suiciderate_newdata.df %>% group_by(Country) %>%
  summarize(suicide_rate=sum(No_of_Suicides)*100000/sum(Population), 
            gdp_per_capita=mean(GDP.per.capita),
            pop=sum(as.numeric(Population)), .groups='drop') %>% 
  arrange(desc(gdp_per_capita)) %>%
  ggplot(aes(gdp_per_capita, suicide_rate)) +
  geom_point(fill="deepskyblue2", color="navy") +
  stat_smooth(method = "lm", color = "red", size = 1) +
  geom_text(data = . %>% filter(gdp_per_capita>64000 | 
            suicide_rate>40), aes(gdp_per_capita, 
            suicide_rate, label=Country, col=Country)) +
  labs(x="Average GDP per capita", y="Suicide rate per 100k population") +
  ggtitle("GDP per Capita vs. Suicide Rate") +
  theme(legend.position = "none") 
```
## Effect of Nation Wealth on Suicide Rates
The next plot shows suicide rates plotted against per capita GDP of the countries in the dataset. It seems that there is a positive linear correlation between suicide rates per 100k population and the country's GDP per capita. There exists some outlier values in the case of Lithuania, where we have relatively small per capita GDP and very high suicide rate (the highest as seen earlier).

```{r}
# Top 10 countries with max suicide rate
max_suicides_plot <- (head(suiciderate_newdata.df %>%
                             group_by(Country) %>% 
                             summarise(population = sum(Population), 
                                       suicides = sum(No_of_Suicides), totalsuiciderate = (suicides / population) * 100000, .groups = 'drop') %>% 
                             arrange(desc(totalsuiciderate)),10) %>% 
                             ggplot(aes(x = Country,y = totalsuiciderate))+ 
                             geom_bar(col = 'skyblue',stat='identity',fill='skyblue')+
                             labs(title = "Total suicidies in top 10 Countries", x = "Countries", y = "Suicides per 100k Population") +
                             theme(axis.text.x = element_text(angle = 90, hjust = 1), axis.text = element_text(size = 10), axis.title = element_text(size =10), plot.title = element_text(size = 10)))
max_suicides_plot
top10countries <- head(suiciderate_newdata.df %>% 
                         group_by(Country) %>% 
                         summarise(population = sum(Population), 
                                   suicides = sum(No_of_Suicides), totalsuiciderate = (suicides / population) * 100000, .groups = 'drop')  %>% 
                         arrange(desc(totalsuiciderate)),10)$Country
```

```{r}
# Age distribution in top 10 countries
country.age.plot <- suiciderate_newdata.df %>% 
  group_by(Country,Age) %>% 
  summarise(suicides_per_100k = mean(Suicide_Per_Population), .groups = 'drop') %>% 
  filter(Country %in% top10countries) %>% 
  ggplot(aes(x = Country, y = suicides_per_100k, fill = factor(Age))) + geom_bar(stat = 'identity',group=1) + labs(title = "Age distribution in top 10 Countries", x = "Countries", y = "Mean  Suicides (per 100k)")   + theme(axis.text.x = element_text(angle = 90, hjust = 1), axis.text = element_text(size = 10), axis.title = element_text(size =10), legend.title = element_text(size =10), legend.text = element_text(size=10), plot.title = element_text(size = 10))
  country.age.plot
```

```{r}
#Trend in top 10 countries by sex
options(repr.plot.width = 10,rer.plot.height = 20)
sex.country.trendplot <- suiciderate_newdata.df %>% 
  group_by(Country,Year,Sex) %>% 
  summarise(suicides_100kpop = mean(Suicide_Per_Population), .groups = 'drop') %>% 
  filter(Country %in% top10countries) %>% 
  ggplot(aes(x = Year,y =suicides_100kpop,col = Sex))+
  geom_line()+geom_point(size = 0.4)+facet_wrap(Country~.,nrow = 5,ncol=2) + scale_x_continuous(breaks = seq(1985, 2015, 5)) + theme(axis.text.x = element_text(angle = 90, hjust =1))
  sex.country.trendplot
```
### Specifying the Model
Since the target variable, `Suicide_Per_Population` is a continuous variable, multiple linear regression algorithm is considered. A regression model is fitted that takes into account continent, population size, per capita CGP, year, sex and age group effects on suicide rates per 100k population.We have taken Continent predictor instead of Country as it is nominal column with too many level. To handle it we added Continent Column to our dataset. The following model is:

```{r}
# Specify explanatory and outcome variables and model formula
vars <- c("Continent", "Population", "Sex", "Year", "Age", "GDP.per.capita")
outcome <- "suicide_rate_log"
(fmla <- as.formula(paste(outcome, "~", paste(vars, collapse = " + "))))
```


### Create Training and Testing sets
Training and testing datasets are created. Testing set is 20% of the entire dataset.

```{r}
# Split to training and testing datasets
set.seed(30)
test_index <- createDataPartition(y = suiciderate_newdata.df$suicide_rate_log, times = 1, 
            p = 0.2, list = FALSE)
train <- suiciderate_newdata.df[-test_index,]
test <- suiciderate_newdata.df[test_index,]

```

### Train the Model
#### Linear Regression
```{r}
# Linear regression
lm1 <- train %>% 
  lm(fmla, data=.)
summary(lm1)
```

#### Random Forests
```{r}
# Random forests
set.seed(30)
rf1 <- ranger(fmla, # formula 
       train, # data
       num.trees = 100, 
       respect.unordered.factors = "order",
       seed = 1)
summary(rf1)
```

### Test the Model
After this, the model is applied to the test data. This is done through the `predict` function. To test the performance of the model, the Root Mean Squared Errors (RMSE) is considered. After generating the model predictions, the RMSE is calculated by comparing the model predictions against the true value of the suicide rates.
```{r}
# Generate predictions using the test data
test$lm <- predict(lm1, newdata = test)
test$rf <- predict(rf1, test)$predictions


# Calculate RMSE
case1 <- test %>% gather(key=model, value=log_pred, lm, rf) %>%
  mutate(pred=exp(log_pred),
         residuals=Suicide_Per_Population-pred) %>%
  group_by(model) %>%
  summarize(rmse=sqrt(mean(residuals^2)), .groups='drop')
```
The model with the lowest RMSE in this case is random forests.

While doing the Exploratory Data Analyisis, we found that the year 2016 has very less data and hence we are going to drop the data for year 2016 and run the regression for the trimmed dataset.

Training and testing data sets are generated again to accommodate the above changes for each case and the models are refitted. Comparison between models performance is done through RMSE values.

```{r}
############  Model Attempt2 ############
suicide_trim <- suiciderate_newdata.df %>% 
  filter(Year!=2016)

# Split to training and testing datasets
set.seed(30)
test_index_trim <- createDataPartition(y = suicide_trim$suicide_rate_log, times = 1, p = 0.2, list = FALSE)
train_trim <- suicide_trim[-test_index_trim,]
test_trim <- suicide_trim[test_index_trim,]

# Train the model
# Linear regression
lm2 <- train_trim %>% 
  lm(fmla, data=.)

# Random forests
set.seed(1, sample.kind="Rounding")
rf2 <- ranger(fmla, # formula 
       train_trim, # data
       num.trees = 100, 
       respect.unordered.factors = "order",
       seed = 1)

# Generate predictions using the test data
test_trim$lm <- predict(lm2, newdata = test_trim)
test_trim$rf <- predict(rf2, test_trim)$predictions

# Calculate RMSE
case2 <- test_trim %>% gather(key=model, value=log_pred, lm, rf) %>%
  mutate(pred=exp(log_pred),
         residuals=Suicide_Per_Population-pred) %>%
  group_by(model) %>%
  summarize(rmse=sqrt(mean(residuals^2)))
```
## 4. Results and Conclusions
The suicide rates per 100k populations prediction algorithm includes several country-level variables in addition to population demographic characteristics. The model used for prediction is a multiple linear regression model fitted to the training data and tested on the testing data. Random Forests model is also attempted at number of trees of 500. Models are fitted based on 2 different cases. Assessment of the perfomance is based on the value of RMSE.

The following table summarizes the performance of each model at the 2 specified cases:

```{r}
case1
case2
```
The Random Forests model yielded the least value of RMSE and is considered the best model for predicting suicide rates per 100k population. Eliminating data of the year 2016 did not improve the value of RMSE.

The following graph shows predictions from the linear regression model and the random forests model compared to the true values of suicide rates in the test dataset for each year. The plot shows that annual predictions generated by random forests model are closer to the true value of suicide rates than those preducted by the linear regression model.

```{r}
test %>% mutate(lm=exp(lm), rf=exp(rf)) %>%
  gather(key=valuetype, value=rate, Suicide_Per_Population, lm, rf) %>%
  mutate(suicides=rate*Population/100000) %>%
  group_by(Year, valuetype) %>%
  mutate(rate_year=sum(suicides)*100000/sum(Population)) %>%
  ggplot(aes(Year, rate_year, col=valuetype)) +
  geom_line() +
  geom_point(size = 2) +
  scale_x_continuous(breaks = seq(1985, 2016, 2)) +
  theme(axis.text.x = element_text(angle = 45))
```

There are other variables that exist in the dataset but not included in the analysis and the modeling. These variables are: 

- `gdp_for_year`, which is the GDP of the country at a given year. It is eliminated as it is highly correlated with `gdp_per_capita` to eliminate multicollinearity in the model. `gdp_per_capita` was selected over `gdp_for_year` as it is a better measure for the GDP and wealth of the nations that takes into account population size.

- `generation`, which is a categorical variable for the generation of the population. It is left out as it is highly correlated with `age`. `age` is selected over `generation` to include in the analysis as it more easily and intuitively understood.

- `HDI.for.year`, as over two thirds of the variable is missing.
